### 2021-06-09 rad-icfp-2021/correctness.tex
some comments
change outline
outline paper in intro
writing about target language
minor paper

\section{Correctness}

%decompose the macro from section 4 into unary form transfo, then principled macro, then opti
\MH{Question: when is it best to give the idea about our macro? that it's UNF+macro+easy-opti?}
\MH{probably won't use UNF in this paper}

\subsection{Unary Form} % (fold)
\label{sub:unary_form}

\subsection{Principled Macro} % (fold)
\label{sub:principled_macro}

\subsection{Correctness via subsconing} % (fold)
\label{sub:correctness_via_subsconing}

\subsection{Optimizations} % (fold)
\label{sub:optimizations}
%the ones in the macro of Sec4 then the further classical ones to ANF

\subsection{Correctness of the macro from section \ref{}} % (fold)
\label{sub:correctness_of_the_macro_from_section_ref}

>>>>

\section{Correctness}
\label{sec:correctness}

We now prove that our transformations are correct. 
The main result of the section is

\begin{theorem}
    For every $\Gamma \vdash$ !e: \reals!, we have 
    $\sem{\pi_2 \invUNF{\UNF{e}}(Id_\Gamma)(last(1:\reals))}= \nabla_{\Gamma}\sem{e}$.
\end{theorem}
where $Id_\Gamma$ is the identity function, 
$\sem{-}$ an appropriate denotational semantics of smooth functions,
and $last(a)$ is the hot-vecteur encoding $(0,\ldots,0,a)$.

\subsection{Denotational semantics}

\begin{itemize}
    \item for Syn
    \item for target
    \item for UNF
\end{itemize}

\MH{find somewhere to put the operational semantics}
\input{figures/op_semantics_target}

\subsection{Correctness}

In order to prove correctness, we need some key intermediate lemmas.

\begin{lemma}
    $\sem{\pi_1\invUNF{\UNF{e}}}=\sem{e}$.
\end{lemma}

\begin{lemma}
    $\sem{\pi_2\invUNF{e}}=T^*\sem{\pi_1\invUNF{e}}$
\end{lemma}

From this, we deduce the main theorem

\begin{theorem}
    \MH{slight generalisation of the corrolary for compositionality reasons}
\end{theorem}

The main theorem specializes to our case of interest

\begin{corollary}
    For every $\Gamma \vdash$ !e: \reals!, we have 
    $\sem{\pi_2 \invUNF{\UNF{e}}(Id_\Gamma)(last(1:\reals))}= \nabla_{\Gamma}\sem{e}$.
\end{corollary}

### 2021-06-09 rad-icfp-2021/intro.tex
minor paper

For example, \MH{TODO: show how $\prod_i v_i$ for a vector $v$ is done in some functional AD systems}.
For real world AD systems, modularity is key. They need to be expressive but have a lot of optimised derivatives for operators that are used a lot.
This is easily seen on numpy \MH{TODO: explain a bit better why torch is good by using optimised grad for numpy.}
An efficient purely functional reverse-mode transformation would be more easily explainable, optimisable, and potentially reusable.

@@

The problems that this raises and that our paper deals with are the following. First, there is a tradeoff between a general expressive language 
as opposed to a domain specific language. The latter usually has more static information and a specific representation to perform optimisations.
Then, a lot of optimisations performed on AD implementations consist of a lot of hand-crafted derivatives for a lot of useful operations like matrix-matrix multiplication, dot-product, etc.
They don't seem to come with more theoretical justification, are error prone, and can hardly fit more general optimisations.
\MH{TODO: maybe the example and ref that ML systems spend 90percent of the time outside of numpy.}
This makes these systems harder to prove correct. The problem is thus to stay provably correct and pure, and yet efficient. What's more is we would like something easily provably efficient. 
A real implementation on top of our work would of course optimise further, potentially using hand-crafted operations as well, but it would be based on more solid grounds.

@@

Our contribution tries to make these claims precise. Our detailed  contribution consist of
\begin{itemize}
    \item introducing an expressive source language with array operations, yet restrictive enough for efficient differentiation
    \item a method for extending our work and modular method for adding and prove correct new primitives as needed in practice 
    \item a new intermediate representation (I.R.) which we call unary form(UNF), essentially turning every operation as a unary one
    \item a simple reverse-mode transformation on the I.R.
    \item an expressive purely functional target language and efficient transformation from the I.R. to this language
    \item several useful generalisations in Section~\ref{sec:generalisation}. 
    \item correctness of the differentiation \ref{sec:correctness}, and complexity guarantees \ref{sec:complexity}
    \item showing how efficiency  is recovered via standard optimisation techniques
    \item a detailed comparison with other work and some limitations to our work in Section~\ref{sec:conclusion}
\end{itemize}

@@

After some general partial evaluation that will be detailed further in the paper, we obtain     
        \begin{center}
            \begin{tabular}{l}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !let y1 = 0 in!\\
                !let y2 = 0 in!\\
                !let y3 = 0 in!\\
                !let y4 = 0 in!\\
                !let y5 = 1 in!\\
                !let y1'! != y1+w1*y5 in!\\
                !let y4'! != y4+x1*y5 in!\\
                !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
            \end{tabular}
        \end{center}  
For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.\\
Of course, this term can be further optimised to give
        \begin{center}
            \begin{tabular}{{c}}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !(w1+x2*x1, x1*x1, 0)!
            \end{tabular}
        \end{center}
    \end{example}

@@

The gradient of each of these construct w.r.t. !A! will be:
\begin{center}
    \begin{tabular}{{r c l}}
        $\nabla_A$!prod(A)! &:=& !map2 * (scanr x 1 A) (shift1(scanl * 1 A))!\\
        $\nabla_A$!sum(A)! &:=& !map (x,y -> 1) A!\\
        $\nabla_A$!dot(A,B)! &:=& !map2 * B (map (x,y -> 1) A! 
    \end{tabular}
\end{center}
where !shift1 [v1,...,vn] = [1,v1,...,v(n-1)]! is a simple operator yet perhaps strange at first naturally showing up in our constructions.
!scanl! and !scanr! are respectively scan left and scan right. These are like fold but store all intermediate results in an array.
We obtain these gradients, which are efficient and usually derived by hand, as special cases of our work.
They are obtained automatically in our case.
\end{example}

>>>>

% For example, \MH{TODO: show how $\prod_i v_i$ for a vector $v$ is done in some functional AD systems}.
% For real world AD systems, modularity is key. They need to be expressive but have a lot of optimised derivatives for operators that are used a lot.
% This is easily seen on numpy \MH{TODO: explain a bit better why torch is good by using optimised grad for numpy.}
An efficient purely functional reverse-mode transformation would be more easily explainable, optimisable, and reusable.

@@

This raises several problems that are dealt with in our paper. First, there is a tradeoff between a general expressive language 
as opposed to a domain specific language. The latter usually has more static information and a specific representation to perform optimisations.
Then, a lot of optimisations performed on AD implementations consist of a lot of hand-crafted derivatives for a lot of useful operations like matrix-matrix multiplication, dot-product, etc.
They don't seem to come with more theoretical justification, are error prone, and can hardly fit more general optimisations.
\MH{TODO: maybe the example and ref that ML systems spend 90percent of the time outside of numpy.}
This makes these systems harder to prove correct. The problem is thus to stay provably correct and pure, and yet efficient. What's more is we would like something easily provably efficient. 
A real implementation on top of our work would of course optimise further, potentially using hand-crafted operations as well, but it would be based on more solid grounds.

@@

Our contribution tries to make these claims precise. Our detailed  contribution consist of
\begin{itemize}
    \item introducing an expressive source language with array operations, yet restrictive enough for efficient differentiation
    \item a method for extending our work and modular method for adding and prove correct new primitives as needed in practice 
    \item a new intermediate representation (I.R.) which we call unary form (UNF), essentially turning every operation to a unary one
    \item a simple reverse-mode transformation on the I.R.
    \item an expressive purely functional target language and efficient transformation from the I.R. to this language
    \item several useful generalisations in Section~\ref{sec:generalisation}. 
    \item correctness of the differentiation \ref{sec:correctness}, and complexity guarantees \ref{sec:complexity}
    \item showing how efficiency  is recovered via standard optimisation techniques
    \item a detailed comparison with other work and some limitations to our work in Section~\ref{sec:conclusion}
\end{itemize}

@@

After some general partial evaluation techniques that will be detailed further in the paper, we obtain     
        \begin{center}
            \begin{tabular}{l}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !let y1 = 0 in!\\
                !let y2 = 0 in!\\
                !let y3 = 0 in!\\
                !let y4 = 0 in!\\
                !let y5 = 1 in!\\
                !let y1'! != y1+w1*y5 in!\\
                !let y4'! != y4+x1*y5 in!\\
                !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
            \end{tabular}
        \end{center}  
For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.\\
Of course, this term can be further optimised via constant propagation and algebraic simplifications to give
        \begin{center}
            \begin{tabular}{{c}}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !(w1+x2*x1, x1*x1, 0)!
            \end{tabular}
        \end{center}
    \end{example}

@@

The gradient of each of these construct w.r.t. !A! will be:
\begin{center}
    \begin{tabular}{{r c l}}
        $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 A) (shift1(scanl * 1 A))!\\
        $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
        $\nabla_A$!dot(A,B)! &:=& !map2 * B (map (x -> 1) A)! 
    \end{tabular}
\end{center}
where !shift1 [v1,...,vn] = [1,v1,...,v(n-1)]! is a simple operator yet perhaps strange at first naturally showing up in our constructions.
!scanl! and !scanr! are respectively scan left and scan right. These are like fold but store all intermediate results in an array.
We obtain these gradients, which are efficient and usually derived by hand, as special cases of our work.
They are obtained automatically in our case.
\end{example}
